`{peacesciencer}`:  A User’s  Guide  for  Quantitative  Peace Science in R

### 2021-06-21 user-guide/ms.Rmd
update the user guide for resubmission

This data feature tackles a recurring problem for researchers in the peace science community. True research reproducibility is achieved creating original dyad-year or state-year data from scratch, though no published guide exists that informs how to do this on their own. Instead, researchers may end up reusing old code generated in past studies, leaving them to spend time and energy adjusting the sample of states, the temporal domain, and doing whatever additional troubleshooting may arise from this practice. In a related manner, researchers may spend too much time reproducing old code for standard information that goes into any dyad-year or state-year analysis---like contiguity relationships and democracy, among many others---and have to do additional troubleshooting for how these various data sources treat missing data or treat state codes in a manner inconsistent with the more accessible Correlates of War or Gleditsch-Ward state codes. This is all compounded by changes in technology that treat the creation of the data and the analysis of data as a continuous process in which the contemporary quantitative political scientist is increasingly becoming a computer programmer as well. Graduate students, in particular, face unique challenges associated with these developments. Graduate students beginning their careers in peace science must learn how scholarship informs data in peace science and how data inform scholarship at the same time they are needing to learn quantitative methods in a chosen software package.

`{peacesciencer}` is designed to address these problems. Built around the free and open source R programming language, `{peacesciencer}` contains a suite of data and functions for creating data of interest to researchers. Researchers can use `{peacesciencer}` to create dyad-year, state-year, and even state-day data from scratch. Afterward, they can add a variety of standard information (e.g. contiguity, alliances, major power status, GDP per capita estimates, capability estimates, and more) to these data with a simple command. This is a considerable time-saver since, in the absence of it, researchers would have to more meticulously code and transform the raw data to conform to the kind of data they want. `{peacesciencer}` comes with some data innovations as well, including a comprehensive data set on democracy by year, an original data set on capitals and capital transitions, and a function to create peace years between ongoing dyad-year or state-year conflicts. All are done with the maximum possible transparency. The project is available for public view of Github (https://github.com/svmiller/peacesciencer). The `data-raw` directory on the project's Github contains information and comments about how every data set was created and the `R` directory informs the user how each function works. The function manuals (`r params$ps_manual`) contain additional comments about what each function returns and, in appropriate cases, why it is doing what it is doing. Thus, `{peacesciencer}` not only assists a peace scientist with their research, but it does so in a manner that best conforms to the  Data Access and Research Transparency Initiative (DA-RT) initiative across all political science.

@@

For one, researchers invest too much time in the construction of a data set that faithfully captures the unit of analysis. Assume a researcher wants an original data set on all directed dyad-years for Correlates of War states for an analysis of inter-state conflict. How might one do that? The answer has never been immediately obvious. No published guide exists that shows a researcher how to create these data themselves from scratch, which is one reason why software bundles like EUGene and NewGene are attractive to researchers who primarily care about the substance of their research question. After all, EUGene's primary impetus was allowing researchers to create data sets for replication of previous studies, producing a host of data types (e.g. dyad-year, state-year, dispute-year) along the way that users can amend as they saw fit. `{peacesciencer}` is primarily born from this question about how to create dyad-year and state-year data from scratch. The underlying code that produces these data types is available online and are converted into simple functions for the ease of the researcher.

@@

Third, the creation of the data and the analysis of the data is increasingly becoming one continuous process. Not too long ago, it used to be the case that researchers had to download a data set, or create one from scratch (possibly in a spreadsheet or through a program like EUGene). After downloading or constructing the data, the researcher then opened a specialty program for statistical analysis (e.g. SAS, SPSS, Stata) to recode raw data into a form suitable for analysis before running a statistical model that regresses some outcome on a set of covariates. Current research practices still resemble this process, but the steps between them are no longer as large as they were in the past. Software options exist that allow the researcher to load data, create data, clean data, analyze data, and present the results of the analysis all within one program. `{peacesciencer}`, by itself, does not do all these things, but it seamlessly connects the beginning of the research process to the end of the research process without needing to leave the increasingly popular R programming language and Rstudio (its free-for-use integrated desktop environment [IDE]).

@@

Fifth, statistical software should be kind to graduate students in peace science with the aforementioned point in mind. Graduate students just beginning their career in peace science scholarship are tasked to learn the substantive material, the characteristics of the underlying data in wide use in the community, and the quantitative methods and statistical software available to them, effectively all at the same time. A student interested in civil conflict scholarship will need to learn the UCDP-PRIO armed conflict data in some detail at the same time they learn about the theoretical arguments that informed the creation of the data, likely at the same time they take quantitative methods courses that show them how to load and explore data in a chosen statistical software. At the same time that creating data is increasingly blurred with coding and processing data, students are simultaneously learning to explore data for themselves as they read about the analyses of the data. `{peacesciencer}` aims to make this process easier for graduate students using the features of the R programming language. Graduate students can use `{peacesciencer}` to learn about peace science data and computer programming in R at the same time.

@@

`{peacesciencer}` comes with a battery of built-in functions for generating the most widespread forms of peace science data. Each function uses raw or pre-processed data included in `{peacesciencer}`. For example, `create_dyadyears()` transforms the raw state system membership data for either the Correlates of War or Gleditsch-Ward system into the full universe of post-1816 dyad-years. `add_gml_mids()` uses a dyadic dispute-year version of the MID data offered by @gibleretal2016amid, which, as the `data-raw` directory on the project's Github shows, is itself derived from the dispute-level and participant-level versions of the data. `{peacesciencer}` also has some data innovations. It has an original data set on capital-to-capital distances, calculating for distance in kilometers between state capitals using the "Vincenty" method (i.e. "as the crow flies") that also accounts for instances when capitals moved (e.g. Brazil in 1960, Burundi in 2018). It also features an innovation of democracy data, providing reasonable estimates of democracy using the @marquez2016qme of extending the Unified Democracy Scores (UDS) data [@pemsteinetal2010dc].[^demvig] As of writing, `{peacesciencer}` comes with the following functionality.

[^demvig]: A vignette on the `{peacesciencer}` website explains just how much unnecessary missingess pervades data on democracy and how this might adversely affect inferences of inter-state or intra-state conflict.

```{r, echo=F, eval=T}
tribble(~data, ~description, ~type, ~func,
        "(CoW, G-W) Dyad-Year Data", "Create dyad-years from state system membership data", "", "create_dyadyears()",
        "(CoW, G-W) State-Year Data", "Create state-years from state system membership data", "", "create_stateyears()",
        "(CoW, G-W) State-Day Data", "Create state-days from state system membership data", "", "create_statedays()",
        "Archigos (v. 4.1)", "Add Archigos political leader information", "D, S", "add_archigos()",
        "ATOP Alliances (v. 5.0)", "Add ATOP alliance data", "D", "add_atop_alliance()",
        "Capital-to-Capital Distances", "Add capital-to-capital distance", "D, S", "add_capital_distance()",
        "Convert CoW codes to G-W codes", "Add matching CoW codes for G-W states by year", "D, S", "add_ccode_to_gw()",
        "Convert G-W codes to CoW codes", "Add matching G-W codes for CoW states by year", "D, S", "add_gwcode_to_cow()",
        "CoW National Capabilities (v. 5.0)", "Adds capabilities of states by year", "D, S", "add_nmc()",
        "CoW Direct Contiguity (v. 3.2)", "Adds contiguity information by year", "D, S", "add_contiguity()",
        "CoW Alliances (v. 4.1)", "Adds CoW alliance data by year", "D", "add_cow_alliance()",
        "CoW IGOs (v. 3.0)", "Adds CoW IGO data by year", "D, S", "add_igos()",
        "CoW Major Powers (v. 2016)", "Adds CoW major power by year", "D, S", "add_cow_majors()",
        "CoW MIDs (v. 5.0)", "Add CoW MID information to data", "D", "add_cow_mids()",
        "CoW Trade Data (v. 4.0)", "Add CoW trade information to data", "D, S", "add_cow_trade()",
        "CoW Wars", "Add CoW inter-state (v. 4.0) and intra-state (v. 4.1) war data", "D", "add_cow_wars()",
        "CREG (v. 1.02)", "Add ethnic/religious fractionalization/polarization by year", "D, S", "add_creg_fractionalization()",
        "Democracy", "Add Polity (v. 2017), V-Dem (v. 10), and Marquez' (2016) UDS extensions", "D,S", "add_democracy()",
        "GML MIDs (v. 2.1.1)", "Add Gibler-Miller-Little MIDs to data", "D", "add_gml_mids()",
        "Minimum Distance (v. 2016)", "Adds minimum distance data by year", "D, S", "add_minimum_distance()",
        "Peace Years", "Calculate peace years for various disputes", "D, S", "add_peace_years()",
        "Rugged/Mountainous Terrain", "Add 'ruggedness'/'mountainous' terrain data by year", "D, S", "add_rugged_terrain()",
        "(Surplus/Gross) Domestic Product", "Add Anders et al. (2020) GDP/SDP simulations by year", "D, S", "add_sdp_gdp()",
        "Strategic Rivalries", "Add Thompson and Dreyer (2012) strategic rivalry data by year", "D, S", "add_strategic_rivalries()",
        "UCDP Armed Conflicts (v. 20.1)", "Add UCDP armed conflict data by year", "S", "add_ucdp_acd()"
        ) -> included

@@

if (is_latex) {
kable(included, booktabs = FALSE, longtable = TRUE,
    caption = "Features of \\{peacesciencer\\}",
    align = c("l","l","c", "c"),
    col.names = c("Data", "Description", "Type", "Function")) %>%
  kable_styling(font_size = 7) %>%
  row_spec(0, bold=TRUE) %>%
  column_spec(4, monospace = TRUE) %>%
  column_spec(c(1,2,3), border_left = FALSE) %>%
  row_spec((head(c(1:nrow(included)), -1)), extra_latex_after = "\\arrayrulecolor{white}") %>%
  row_spec((included %>% 
             nrow()), extra_latex_after = "\\arrayrulecolor{black}")
} else {
  kable(included,
    caption = "Features of \\{peacesciencer\\}",
    align = c("l","l","c", "c"),
    col.names = c("Data", "Description", "Type", "Function")) %>%
  # kable_styling(font_size = 7) %>%
  row_spec(0, bold=TRUE) %>%
  column_spec(4, monospace = TRUE) 
}

@@

For example, assume a researcher wanted to replicate something analogous to Bremer's [-@bremer1992ddc] famous "dangerous dyads" analysis about the causes of conflict onset. `{peacesciencer}` can do something like this in the following pipe-based workflow. First, `create_dyadyears(directed = FALSE, mry = FALSE)` will create a non-directed dyad-year data frame for all Correlates of War states from 1816 to 2016.

@@

[^onpy]: `add_peace_years()` will only calculate the peace years---titled, in this case, `gmlmidspell`---and will leave the temporal dependence adjustment to the taste of the researcher. Importantly, I do not recommend manually creating splines or square/cube terms because it creates more problems in adjusting for temporal dependence in model predictions. In a regression formula in R, you can specify the @cartersignorino2010bf approach as `... + gmlmidspell + I(gmlmidspell^2) + I(gmlmidspell^3)`. The @becketal1998tts cubic splines approach is `... + splines::bs(gmlmidspell, 4)`. This function includes the spell and three splines (hence the 4 in the command).

@@

```{r, echo=T, eval=F}
Data %>%
  mutate(landcontig = ifelse(conttype == 1, 1, 0)) %>%
  mutate(cowmajdyad = ifelse(cowmaj1 == 1 | cowmaj2 == 1, 1, 0)) %>%
  # Create estimate of militarization as milper/tpop
  # Then make a weak-link
  mutate(milit1 = milper1/tpop1,
         milit2 = milper2/tpop2,
         minmilit = ifelse(milit1 > milit2,
                           milit2, milit1)) %>%
  # create CINC proportion (lower over higher)
  mutate(cincprop = ifelse(cinc1 > cinc2,
                           cinc2/cinc1, cinc1/cinc2)) %>%
  # create weak-link specification using Quick UDS data
  mutate(mindemest = ifelse(xm_qudsest1 > xm_qudsest2,
                            xm_qudsest2, xm_qudsest1)) %>%
  # Create "weak-link" measure of jointly advanced economies
  mutate(minwbgdppc = ifelse(wbgdppc2011est1 > wbgdppc2011est2,
                             wbgdppc2011est2, wbgdppc2011est1)) -> Data

>>>>

This data feature tackles a recurring problem for researchers in the peace science community. True research reproducibility is achieved creating original dyad-year or state-year data from scratch, though no published guide exists that informs researchers how to do this on their own. Instead, researchers may end up reusing old code generated in past studies, leaving them to spend time and energy adjusting the sample of states, the temporal domain, and doing whatever additional troubleshooting may arise from this practice. Researchers may additionally spend too much time reproducing old code for standard information that goes into any dyad-year or state-year analysis---like contiguity relationships and democracy---and have to do additional troubleshooting for how these various data sources treat missing data or treat state codes in a manner inconsistent with the more accessible Correlates of War or Gleditsch-Ward state codes. This is all compounded by changes in technology that treat the creation of the data and the analysis of data as a continuous process in which the contemporary quantitative political scientist is increasingly becoming a computer programmer as well. Graduate students, in particular, face unique challenges associated with these developments. Graduate students beginning their careers in peace science must learn how scholarship informs data in peace science and how data inform scholarship at the same time they are needing to learn quantitative methods in a chosen software package.

`{peacesciencer}` is designed to address these problems. Built around the free and open source R programming language, `{peacesciencer}` contains a suite of data and functions for creating data of interest to researchers. Researchers can use `{peacesciencer}` to create dyad-year, state-year, and even state-day data from scratch. Afterward, they can add a variety of standard information (e.g. contiguity, alliances, major power status, GDP per capita estimates, capability estimates, and more) to these data with a simple command. This is a considerable time-saver since, in the absence of it, researchers would have to more meticulously code and transform the raw data to conform to the kind of data they want. `{peacesciencer}` comes with some data innovations as well, including a comprehensive data set on democracy by year, an original data set on capitals and capital transitions, and a function to create peace years between ongoing dyad-year or state-year conflicts. All are done with the maximum possible transparency. The project is available for public view of Github (`r params$ps_github`). The `data-raw` directory on the project's Github contains information and comments about how every data set was created and the `R` directory informs the user how each function works. The function manuals (`r params$ps_manual`) contain additional comments about what each function returns and, in appropriate cases, why it is doing what it is doing. Thus, `{peacesciencer}` not only assists a peace scientist with their research, but it does so in a manner that best conforms to the  Data Access and Research Transparency Initiative (DA-RT) initiative across all political science.

@@

For one, researchers invest too much time in the construction of a data set that faithfully captures the unit of analysis. Assume a researcher wants an original data set on all directed dyad-years for Correlates of War states for an analysis of inter-state conflict. How might one do that? The answer has never been immediately obvious. No published guide exists that shows a researcher how to create these data themselves from scratch, which is one reason why software bundles like EUGene and NewGene are attractive to researchers who primarily care about the substance of their research question. After all, EUGene's primary impetus was allowing researchers to create data sets for replication of previous studies, producing a host of data types (e.g. dyad-year, state-year, dispute-year) along the way that users can amend as they saw fit. `{peacesciencer}` is primarily born from this question about how to create dyad-year and state-year data from scratch. The underlying code that produces these data types is available online and `{peacesciencer}` converts these lines of code into simple functions for the ease of the researcher.

@@

Third, the creation of the data and the analysis of the data are increasingly becoming one continuous process. Not too long ago, it used to be the case that researchers had to download a data set, or create one from scratch (possibly in a spreadsheet or through a program like EUGene). After downloading or constructing the data, the researcher then opened a specialty program for statistical analysis (e.g. SAS, SPSS, Stata) to recode raw data into a form suitable for analysis before running a statistical model that regresses some outcome on a set of covariates. Current research practices still resemble this process, but the steps between them are no longer as large as they were in the past. Software options exist that allow the researcher to load data, create data, clean data, analyze data, and present the results of the analysis all within one program. `{peacesciencer}`, by itself, does not do all these things, but it seamlessly connects the beginning of the research process to the end of the research process without needing to leave the increasingly popular R programming language and Rstudio (its free-for-use integrated desktop environment [IDE]).

@@

Fifth, statistical software should be kind to graduate students in peace science with the aforementioned points in mind. Graduate students just beginning their career in peace science scholarship are tasked to learn the substantive material, the characteristics of the underlying data in wide use in the community, and the quantitative methods and statistical software available to them, effectively all at the same time. A student interested in civil conflict scholarship will need to learn the UCDP-PRIO armed conflict data in some detail at the same time they learn about the theoretical arguments that informed the creation of the data, likely at the same time they take quantitative methods courses that show them how to load and explore data in a chosen statistical software. At the same time that creating data is increasingly blurred with coding and processing data, students are simultaneously learning to explore data for themselves as they read about the analyses of the data. `{peacesciencer}` aims to make this process easier for graduate students using the features of the R programming language. Graduate students can use `{peacesciencer}` to learn about peace science data and computer programming in R at the same time.

@@

`{peacesciencer}` comes with a fully developed suite of built-in functions for generating the most widespread forms of peace science data. Each function uses raw or pre-processed data included in `{peacesciencer}`. For example, `create_dyadyears()` transforms the raw state system membership data for either the Correlates of War or Gleditsch-Ward system into the full universe of post-1816 dyad-years. `add_gml_mids()` uses a dyadic dispute-year version of the MID data offered by @gibleretal2016amid, which, as the `data-raw` directory on the project's Github shows, is itself derived from the dispute-level and participant-level versions of the data. `{peacesciencer}` also has some data innovations. It has an original data set on capital-to-capital distances, calculating for distance in kilometers between state capitals using the "Vincenty" method (i.e. "as the crow flies") that also accounts for instances when capitals moved (e.g. Brazil in 1960, Burundi in 2018). It also features an innovation of democracy data, providing reasonable estimates of democracy using the @marquez2016qme method of extending the Unified Democracy Scores (UDS) data [@pemsteinetal2010dc].[^demvig] As of writing, `{peacesciencer}` comes with the following functionality described in Table \@ref(tab:tab-functionality).

[^demvig]: `{peacesciencer}` comes with a vignette that explains just how much unnecessary missingess pervades data on democracy and how this might adversely affect inferences of inter-state or intra-state conflict.

```{r tab-functionality, echo=F, eval=T}
tribble(~data, ~description, ~type, ~func,
        "(CoW, G-W) Dyad-Year Data", "Create dyad-years from state system membership data", "", "create_dyadyears()",
        "(CoW, G-W) State-Year Data", "Create state-years from state system membership data", "", "create_stateyears()",
        "(CoW, G-W) State-Day Data", "Create state-days from state system membership data", "", "create_statedays()",
        "Archigos (v. 4.1)", "Add Archigos political leader information", "D, S", "add_archigos()",
        "ATOP Alliances (v. 5.0)", "Add ATOP alliance data", "D", "add_atop_alliance()",
        "Capital-to-Capital Distances", "Add capital-to-capital distance", "D, S", "add_capital_distance()",
        "Convert CoW codes to G-W codes", "Add matching CoW codes for G-W states by year", "D, S", "add_ccode_to_gw()",
        "Convert G-W codes to CoW codes", "Add matching G-W codes for CoW states by year", "D, S", "add_gwcode_to_cow()",
        "CoW National Capabilities (v. 5.0)", "Adds capabilities of states by year", "D, S", "add_nmc()",
        "CoW Direct Contiguity (v. 3.2)", "Adds contiguity information by year", "D, S", "add_contiguity()",
        "CoW Alliances (v. 4.1)", "Adds CoW alliance data by year", "D", "add_cow_alliance()",
        "CoW IGOs (v. 3.0)", "Adds CoW IGO data by year", "D, S", "add_igos()",
        "CoW Major Powers (v. 2016)", "Adds CoW major power by year", "D, S", "add_cow_majors()",
        "CoW MIDs (v. 5.0)", "Add CoW MID information to data", "D", "add_cow_mids()",
        "CoW Trade Data (v. 4.0)", "Add CoW trade information to data", "D, S", "add_cow_trade()",
        "CoW Wars", "Add CoW inter-state (v. 4.0) and intra-state (v. 4.1) war data", "D", "add_cow_wars()",
        "CREG (v. 1.02)", "Add ethnic/religious fractionalization/polarization by year", "D, S", "add_creg_fractionalization()",
        "Democracy", "Add Polity (v. 2017), V-Dem (v. 10), and Marquez' (2016) UDS extensions", "D,S", "add_democracy()",
        "GML MIDs (v. 2.1.1)", "Add Gibler-Miller-Little MIDs to data", "D", "add_gml_mids()",
        "Minimum Distance (v. 2016)", "Adds minimum distance data by year", "D, S", "add_minimum_distance()",
        "Peace Years", "Calculate peace years for various disputes", "D, S", "add_peace_years()",
        "Rugged/Mountainous Terrain", "Add 'ruggedness'/'mountainous' terrain data by year", "D, S", "add_rugged_terrain()",
        "(Surplus/Gross) Domestic Product", "Add Anders et al. (2020) GDP/SDP simulations by year", "D, S", "add_sdp_gdp()",
        "Strategic Rivalries", "Add Thompson and Dreyer (2012) strategic rivalry data by year", "D, S", "add_strategic_rivalries()",
        "UCDP Armed Conflicts (v. 20.1)", "Add UCDP armed conflict data by year", "S", "add_ucdp_acd()"
        ) -> included

@@

if (is_latex) {
kable(included, #booktabs = FALSE, longtable = TRUE,
    caption = "Features of \\{peacesciencer\\}",
    align = c("l","l","c", "c"),
    col.names = c("Data", "Description", "Type", "Function")) %>%
  kable_styling(font_size = 8) %>%
  row_spec(0, bold=TRUE) %>%
  column_spec(4, monospace = TRUE) %>%
  column_spec(c(1,2,3), border_left = FALSE) %>%
  row_spec((head(c(1:nrow(included)), -1)), extra_latex_after = "\\arrayrulecolor{white}") %>%
  row_spec((included %>% 
             nrow()), extra_latex_after = "\\arrayrulecolor{black}")
} else {
  kable(included,
    caption = "Features of \\{peacesciencer\\}",
    align = c("l","l","c", "c"),
    col.names = c("Data", "Description", "Type", "Function")) %>%
  # kable_styling(font_size = 7) %>%
  row_spec(0, bold=TRUE) %>%
  column_spec(4, monospace = TRUE) 
}

@@

For example, assume a researcher wanted to replicate something analogous to Bremer's [-@bremer1992ddc] famous "dangerous dyads" analysis about the causes of conflict onset. `{peacesciencer}` can do something like this in the following pipe-based workflow. First, `create_dyadyears(directed = FALSE, mry = FALSE)` will create a non-directed dyad-year data frame for all Correlates of War states from 1816 to 2016.[^after2010]

[^after2010]: The conflict data in @gibleretal2016amid do not extend past 2010, but asking for dyad-years from 2010-2016 will just create harmless `NA`s that will not appear in any analysis.

@@

[^onpy]: `add_peace_years()` will only calculate the peace years---titled, in this case, `gmlmidspell`---and will leave the temporal dependence adjustment to the taste of the researcher. Importantly, I do not recommend manually creating splines or square/cube terms because it creates more problems in adjusting for temporal dependence in model predictions. In a regression formula in R, you can specify the @cartersignorino2010bf approach as `... + gmlmidspell + I(gmlmidspell^2) + I(gmlmidspell^3)`. The @becketal1998tts cubic splines approach is `... + splines::bs(gmlmidspell, 4)`. This function includes the spell and three splines (hence the 4 in the command). Either approach makes for easier model predictions, given R's functionality.

@@

```{r, echo=T, eval=F}
Data %>%
  mutate(landcontig = ifelse(conttype == 1, 1, 0)) %>%
  mutate(cowmajdyad = ifelse(cowmaj1 == 1 | cowmaj2 == 1, 1, 0)) %>%
  # Create estimate of militarization as milper/tpop
  # Then make a weak-link
  mutate(milit1 = milper1/tpop1,
         milit2 = milper2/tpop2,
         minmilit = ifelse(milit1 > milit2,
                           milit2, milit1)) %>%
  # create CINC proportion (lower over higher)
  mutate(cincprop = ifelse(cinc1 > cinc2,
                           cinc2/cinc1, cinc1/cinc2)) %>%
  # create weak-link specification using Quick UDS data
  mutate(mindemest = ifelse(xm_qudsest1 > xm_qudsest2,
                            xm_qudsest2, xm_qudsest1)) %>%
  # Create "weak-link" measure of jointly advanced economies
  mutate(minwbgdppc = ifelse(wbgdppc2011est1 > wbgdppc2011est2,
                             wbgdppc2011est2, wbgdppc2011est1)) -> Data
```
